Miniproject - Elliptic Curves

### 2022-03-28 Elliptic Curves/submission.tex
Added new possibly unnecessary proposition

\section{Question 3}
First, we first look at some basic transformations we can do to Weierstraß equations.
This will be our main toolkit for this exercise.
\begin{prop}[Weierstraß transformations]
    \label{prop:weierstrass_transformations}
    Let $E: y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2 + a_4 x + a_6$ be an Elliptic Curve defined over $k$.
    There are three nice types of transformations
    \begin{description}
        \item[Translation] Let $P = (s, t) \in E(k)$ be a point. Then the isomorphism
        \begin{equation*}
            E \to E', \quad (x, y) \mapsto (x - s, y - t)
        \end{equation*}
        maps $P$ to $(0, 0)$ and the curve $E$ to an Elliptic Curve
        \begin{equation*}
            E': y^2 + a_1 x y + (a_3 + 2t + a_1 s) y = x^3 + (a_2 + 3 s) x^2 + (a_4 - 3 s^2 + 2 a_2 s - a_1 t) x
        \end{equation*}
        This is very useful to clear $a_6$ and continue working with the point $(0, 0)$.
        \item[Shearing] Let $r \in k$. Then the isomorphism
        \begin{equation*}
            E \to E', \quad (x, y) \mapsto (x, y - r x)
        \end{equation*}
        preserves $(0, 0)$ and maps $E$ to an Elliptic Curve
        \begin{equation*}
            E': y^2 + (a_1 + 2r) x y + a_3 y = x^3 + (a_2 + r^2 - a_1 r) x^2 + (a_4 - r a_3) x + a_6
        \end{equation*}
        This is very useful, as it does not change $a_3$ and $a_6$.
        \item[Scaling] Let $u \in k^*$. Then the isomorphism
        \begin{equation*}
            E \to E', \quad (x, y) \mapsto (u^2 x, y^3 y)
        \end{equation*}
        preserves $(0, 0)$ and maps $E$ to an Elliptic Curve
        \begin{equation*}
            E': y^2 + \frac {a_1} u x y + \frac {a_3} {u^3} = x^3 + \frac {a_2} {u^2} x^2 + \frac {a_4} {u^4} x + \frac {a_6} {u^6}
        \end{equation*}
        This is very useful, as it does not change which of the $a_1, ..., a_4, a_6$ are zero.
    \end{description}
\end{prop}
\begin{proof}
    Just plug the equation of the isomorphism into the equation defining the $E'$, and check that it is zero modulo the equation of $E$.
\end{proof}
\begin{corollary}
    \label{prop:point_normalization}
    Let $E$ be an Elliptic Curve defined over $k$ with a $k$-rational point $P$ that is not a 2-torsion point.
    Then there is a linear isomorphism $\psi: E \to E'$ mapping $E$ to an Elliptic Curve of the form
    \begin{equation*}
        E': y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2
    \end{equation*}
    such that $\psi(P) = (0, 0)$, the tangent at $(0, 0)$ is given by $y = 0$.
    Furthermore $[2](0, 0) = (-a_2, a_1 a_2 - a_3)$.
\end{corollary}
\begin{proof}
    After a translation by $-P$, we may assume that
    \begin{equation*}
        E: y^2 + a'_1 x y + a'_3 y = x^3 + a'_2 x^2 + a'_4 x
    \end{equation*}
    and $P = (0, 0)$.
    Now observe that if $a'_3 = 0$, the line $x = 0$ through $(0, 0)$ and $\O$ meets $E$ in $(0, 0)$ with multiplicity 2, and so $(0, 0) + \O = (0, 0)$, contradicting the assumption that $P$ is not a 2-torsion point.
    Thus $a'_3 \neq 0$ and a shearing with $r = a'_4 / a'_3$ maps $E$ to
    \begin{equation*}
        E': y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2
    \end{equation*}
    Note that the tangent at $(0, 0)$ now has slope $0$, i.e. is given by $y = 0$.
    Furthermore, the third point of intersection of the tangent and $E$ is $(-a_2, 0)$.
    The line through $\O$ and $(-a_2)$ is given by $x = -a_2$ and its third point of intersection with $E$ is then $(-a_2, a_1a_2 - a_3)$. 
\end{proof}
I came up with the next lemma to make my first proof of 3(i) work.
Since then, I have found a simpler proof that does not require the lemma anymore, but I found it beautiful and did not want to delete it.
\begin{lemma}
    \label{prop:share_subgroup_equal}
    Let $E, E'$ be Elliptic Curves defined over any field $k$, and assume they share a cyclic subgroup of order $n \geq 5$\footnote{Technically, we can also allow infinite order here.}.
    With this, we mean there is a point $P \in E \cap E'$ of order $n$ such that
    \begin{equation*}
        G := \langle P \rangle_E \subseteq E' \quad \text{and} \quad \restr{+_E}{G} = \restr{+_{E'}}{G}
    \end{equation*}
    Then $E = E'$ (in the sense that they have the same defining equation).
\end{lemma}
\begin{proof}
    Consider some point $[i] P = (a, b) \in E \cap E'$.
    With $P' = (a', b') := -[2i] P \in E \cap E'$ we see that $P + P + P' = \O$ and so $P, P, P'$ are colinear\footnote{We mean that the line through $P$ and $P'$ meets $E$ resp. $E'$ at $P$ with multiplicity 2.}.
    In particular, the tangent on $E$ resp. on $E'$ at $P \in E \cap E'$ both have the slope $(b - b')/(a - a')$\footnote{Or infinity if $a = a'$, but importantly, the slopes are equal.}.

@@

    Now consider the line through $\O$ and $(0, 0)$, which is given by $x = 0$.
    Plugging this in yields $y^2 + v y = 0$ and so the third point of intersection is $(0, -v)$.
    Now note that $(0, 0)$, $(0, -v)$ and $\O$ are colinear, so $(0, 0) + (0, -v) + \O = \O$, hence $(0, 0) = -[2](0, 0)$ has order 3.
\end{proof}

\section{Question 4}

Let $S = \{ x^2 \ | \ x \in \Q^* \}$.
\begin{example}[4(i)]
    \label{ex:4i}
    The Elliptic Curve $E: y^2 = x(x + 6x + 1)$ has rank $0$.
\end{example}
\begin{proof}
    As in the lecture, consider
    \begin{align*}
        &E': y^2 = x(x^2 - 12x + 32) \\
        &\phi: E \to E', \quad (u, v) \mapsto \left( \frac {y^2} {x^2}, \ y \frac {x^2 - 1} {x^2} \right) \\
        &\hat{\phi}: E' \to E, \quad (u, v) \mapsto \left( \frac {y^2} {4x^2}, \ y \frac {x^2 - 1} {8x^2} \right) \\
        &q: E'(\Q)/\phi(E(\Q)) \to \Q^*/S, \quad \overline{(u, v)} \mapsto \begin{cases}
            \overline{u} & \text{if $u \neq 0$} \\
            \overline{32} & \text{if $u = 0$}
        \end{cases} \\
        &\hat{q}: E(\Q)/\phi(E'(\Q)) \to \Q^*/S, \quad \overline{(u, v)} \mapsto \begin{cases}
            \overline{u} & \text{if $u \neq 0$} \\
            \overline{1} & \text{if $u = 0$}
        \end{cases}
    \end{align*}
    \paragraph{Find $E'(\Q)/\phi(E(\Q))$} As in the lecture, consider $r \divides 32$ square-free, i.e. $r \in \{ \pm 1, \pm 2 \}$.

>>>>

\section{Question 3}
First, we first look at some basic transformations we can do to Weierstraß equations.
This will be our main toolkit for this exercise.
\begin{prop}[Weierstraß transformations]
    \label{prop:weierstrass_transformations}
    Let $E: y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2 + a_4 x + a_6$ be an Elliptic Curve defined over $k$.
    There are three nice types of transformations
    \begin{description}
        \item[Translation] Let $P = (s, t) \in E(k)$ be a point. Then the isomorphism
        \begin{equation*}
            E \to E', \quad (x, y) \mapsto (x - s, y - t)
        \end{equation*}
        maps $P$ to $(0, 0)$ and the curve $E$ to an Elliptic Curve
        \begin{equation*}
            E': y^2 + a_1 x y + (a_3 + 2t + a_1 s) y = x^3 + (a_2 + 3 s) x^2 + (a_4 - 3 s^2 + 2 a_2 s - a_1 t) x
        \end{equation*}
        This is very useful to clear $a_6$ and continue working with the point $(0, 0)$.
        \item[Shearing] Let $r \in k$. Then the isomorphism
        \begin{equation*}
            E \to E', \quad (x, y) \mapsto (x, y - r x)
        \end{equation*}
        preserves $(0, 0)$ and maps $E$ to an Elliptic Curve
        \begin{equation*}
            E': y^2 + (a_1 + 2r) x y + a_3 y = x^3 + (a_2 + r^2 - a_1 r) x^2 + (a_4 - r a_3) x + a_6
        \end{equation*}
        This is very useful, as it does not change $a_3$ and $a_6$.
        \item[Scaling] Let $u \in k^*$. Then the isomorphism
        \begin{equation*}
            E \to E', \quad (x, y) \mapsto (u^2 x, y^3 y)
        \end{equation*}
        preserves $(0, 0)$ and maps $E$ to an Elliptic Curve
        \begin{equation*}
            E': y^2 + \frac {a_1} u x y + \frac {a_3} {u^3} = x^3 + \frac {a_2} {u^2} x^2 + \frac {a_4} {u^4} x + \frac {a_6} {u^6}
        \end{equation*}
        This is very useful, as it does not change which of the $a_1, ..., a_4, a_6$ are zero.
    \end{description}
\end{prop}
\begin{proof}
    Just plug the equation of the isomorphism into the equation defining the $E'$, and check that it is zero modulo the equation of $E$.
\end{proof}
\begin{corollary}
    \label{prop:point_normalization}
    Let $E$ be an Elliptic Curve defined over $k$ with a $k$-rational point $P$ that is not a 2-torsion point.
    Then there is an Elliptic Curve $E'$ and a linear isomorphism $\psi: E \to E'$ such that $P \mapsto (0, 0)$ and the tangent at $(0, 0)$ on $E'$ is given by the equation $y = 0$.

    Furthermore $E'$ is given by an equation of the form
    \begin{equation*}
        E': y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2
    \end{equation*}
    and $[2](0, 0) = (-a_2, a_1 a_2 - a_3)$.
\end{corollary}
\begin{proof}
    After a translation by $-P$, we may assume that
    \begin{equation*}
        E: y^2 + a'_1 x y + a'_3 y = x^3 + a'_2 x^2 + a'_4 x
    \end{equation*}
    and $P = (0, 0)$.
    Now observe that if $a'_3 = 0$, the line $x = 0$ through $(0, 0)$ and $\O$ meets $E$ in $(0, 0)$ with multiplicity 2, and so $(0, 0) + \O = (0, 0)$, contradicting the assumption that $P$ is not a 2-torsion point.
    Thus $a'_3 \neq 0$ and a shearing with $r = a'_4 / a'_3$ maps $E$ to
    \begin{equation*}
        E': y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2
    \end{equation*}
    Note that the tangent at $(0, 0)$ now has slope $0$, i.e. is given by $y = 0$.
    Furthermore, the third point of intersection of the tangent and $E$ is $(-a_2, 0)$.
    The line through $\O$ and $(-a_2)$ is given by $x = -a_2$ and its third point of intersection with $E$ is then $(-a_2, a_1a_2 - a_3)$. 
\end{proof}
I came up with the next lemma to make my first proof of 3(i) work.
Since then, I have found a simpler proof that does not require the lemma anymore, but I found it beautiful and did not want to delete it.
\begin{lemma}
    \label{prop:share_subgroup_equal}
    Let $E, E'$ be Elliptic Curves defined over any field $k$, and assume they share a cyclic subgroup of order $n \geq 5$\footnote{Technically, we can also allow infinite order here.}.
    With this, we mean there is a point $P \in E \cap E'$ of order $n$ such that
    \begin{equation*}
        G := \langle P \rangle_E \subseteq E' \quad \text{and} \quad \restr{+_E}{G} = \restr{+_{E'}}{G}
    \end{equation*}
    Then $E = E'$ (in the sense that they have the same defining equation).
\end{lemma}
\begin{proof}
    Consider some point $[i] P = (a, b) \in E \cap E'$.
    With $P' = (a', b') := -[2i] P \in E \cap E'$ we see that $P + P + P' = \O$ and so $P, P, P'$ are colinear\footnote{We mean that the line through $P$ and $P'$ meets $E$ resp. $E'$ at $P$ with multiplicity 2.}.
    In particular, the tangent on $E$ resp. on $E'$ at $P \in E \cap E'$ both have the slope $(b - b')/(a - a')$\footnote{Or infinity if $a = a'$, but importantly, the slopes are equal.}.

@@

    Now consider the line through $\O$ and $(0, 0)$, which is given by $x = 0$.
    Plugging this in yields $y^2 + v y = 0$ and so the third point of intersection is $(0, -v)$.
    Now note that $(0, 0)$, $(0, -v)$ and $\O$ are colinear, so $(0, 0) + (0, -v) + \O = \O$, hence $(0, 0) = -[2](0, 0)$ has order 3.
\end{proof}
A similar approach works also for 4-torsion points.
\begin{prop}(3(ii) - 4-torsion points)
    Let $E$ be an Elliptic Curve defined over $\Q$.
    Then $E(\Q)$ has a nontrivial 4-torsion point if and only if $E$ is isomorphic (over $\Q$) to an Elliptic Curve given by an equation of the form
    \begin{equation*}
        y^2 + x y + v y = x^3 + v x^2
    \end{equation*}
\end{prop}
\begin{proof}
    Again, by Corollary~\ref{prop:point_normalization}, assume wlog that
    \begin{equation*}
        E: y^2 + a_1 x y + a_3 y = x^3 + a_2 x^2
    \end{equation*}
    and $(0, 0) \in E(\Q)$ is a nontrivial 4-torsion point.
    Find that $[2](0, 0) = (-a_2, \beta)$ where $\beta = a_1 a_2 - a_3$.
    The tangent at $(-a_2, \beta)$ must have the equation $x = -a_2$ since $(-a_2, \beta)$ is 2-torsion by assumption.
    Thus
    \begin{equation*}
        0 = 2\beta - a_1a_2 + a_3 = \beta
    \end{equation*}
    and so $a_1a_2 = a_3$.
    By scaling with $a_1$ (which is nonzero, otherwise $a_3 = 0$ and the curve is singular), observe that $E$ is isomorphic to the curve
    \begin{equation*}
        E': y^2 + x y + v y = x^3 + v x^2
    \end{equation*}
    where $v = a_3 / a_1^3 = a_2 / a_1^2$.

    Conversely, assume that $E$ is isomorphic to an Elliptic Curve of the above form, so wlog
    \begin{equation*}
        E: y^2 + x y + v y = x^3 + v x^2
    \end{equation*}
    We show that $(0, 0) \in E(\Q)$ has order $4$.
    The tangent has slope $0$, so is given by the line $y = 0$.
    The third point of intersection with $E$ is now $(-v, 0)$.
    Note that the line through $(-v, 0)$ and $\O$ has the equation $x = -v$ and meets $E$ at $(-v, 0)$ with multiplicity 2.
    It follows that $(-v, 0)$ is a 2-torsion point, and so $E(\Q)$ has the point $(0, 0)$ of order 4.
\end{proof}
Now we strengthen Corollary~\ref{prop:point_normalization} somewhat more.
\begin{lemma}
    Let $E$ be an Elliptic Curve defined over $\Q$ with a $\Q$-rational point $P$ that is not a 2-torsion point.
    Then there exists a unique Elliptic Curve $E'$ and a linear isomorphism\footnote{the isomorphism is only unique up to composition with an automorphism of $E'$.} $\psi: E \to E'$ such that $P \mapsto (0, 0)$ and $E'$ is given by an equation of the form
    \begin{equation*}
        E': y^2 + x y + a_3 y = x^3 + a_2 x^2
    \end{equation*}

    In this case have that the tangent at $(0, 0)$ is given by $y = 0$ and $[2](0, 0) = (-a_2, a_1 a_2 - a_3)$.
\end{lemma}
\begin{proof}
    Existence holds by Corollary~\ref{prop:point_normalization} and applying a scaling by $u = a_1$ (note that this is $\neq 0$, otherwise the curve would be singular).

    So we have to show uniqueness.
    Assume there is a second curve
    \begin{equation*}
        E'': y^2 + x y + a'_3 = x^3 + a'_2 x^2
    \end{equation*}
    and a linear isomorphism $\psi': E \to E''$ mapping $P \to (0, 0)$.
    Then we find a linear isomorphism
    \begin{equation*}
        E' \to E'', \quad (x, y) \mapsto (u^2 x + r, u^3 y + u^2 s x + t)
    \end{equation*}
    Note that it must map $(0, 0) \mapsto (0, 0)$, thus $r = t = 0$.
    Plugging this into the equation of $E''$, we get
    \begin{equation*}
        G := -u^6 x^3 + u^4 s^2 x^2 + 2 u^5 s x y + u^6 y^2 + u^4 s x^2 - u^4 a'_2 x^2 + u^5 x y + u^2 a'_3 s x + u^3 a'_3 y
    \end{equation*}
    This must be zero modulo $F = y^2 + x y + a_3 y - x^3 - a_2 x^2$.
    By a simple degree argument, observe that $G - u^6F$ must already be zero in $\Q[x, y]$.
    Grouping coefficients of $G - u^6F$, we see that
    \begin{align*}
        &u^2 a'_3 s = 0 \\
        &u^6 a_2 + u^4 s^2 + u^4 s - u^4 a'_2 = 0 \\
        -&u^6 a_3 + u^3 a'_3 = 0 \\
        -&u^6 + 2 u^5 s + u^5 = 0
    \end{align*}
    Note that $u \neq 0$, so $a'_3 s = 0$.
    Note that $s \neq 0$, otherwise the isomorphism is just a scaling, and in this case, it must already be the identity.
    Thus $s = 0$ and we find
    \begin{align*}
        &u^2 a_2 - a'_2 = 0 \\
        -&u^3 a_3 + a'_3 = 0 \\
        -&u + 1 = 0
    \end{align*}
    So $u = 1$ and $a_2 = a'_2$, $a_3 = a'_3$.
    This shows that $E' = E''$ and then the isomorphisms $\psi, \psi'$ must be equal up to composition with an automorphism in $E'$.
\end{proof}

\section{Question 4}
Let $S = \{ x^2 \ | \ x \in \Q^* \}$.
\begin{example}[4(i)]
    \label{ex:4i}
    The Elliptic Curve $E: y^2 = x(x + 6x + 1)$ has rank $0$.
\end{example}
\begin{proof}
    As in the lecture, consider
    \begin{align*}
        &E': y^2 = x(x^2 - 12x + 32) \\
        &\phi: E \to E', \quad (u, v) \mapsto \left( \frac {y^2} {x^2}, \ y \frac {x^2 - 1} {x^2} \right) \\
        &\hat{\phi}: E' \to E, \quad (u, v) \mapsto \left( \frac {y^2} {4x^2}, \ y \frac {x^2 - 1} {8x^2} \right) \\
        &q: E'(\Q)/\phi(E(\Q)) \to \Q^*/S, \quad \overline{(u, v)} \mapsto \begin{cases}
            \overline{u} & \text{if $u \neq 0$} \\
            \overline{32} & \text{if $u = 0$}
        \end{cases} \\
        &\hat{q}: E(\Q)/\phi(E'(\Q)) \to \Q^*/S, \quad \overline{(u, v)} \mapsto \begin{cases}
            \overline{u} & \text{if $u \neq 0$} \\
            \overline{1} & \text{if $u = 0$}
        \end{cases}
    \end{align*}
    \paragraph{Find $E'(\Q)/\phi(E(\Q))$} As in the lecture, consider $r \divides 32$ square-free, i.e. $r \in \{ \pm 1, \pm 2 \}$.
